Visual Odometry for Autonomous Underwater Vehicles

### 2013-04-04 paper.tex
Small revisions from stephan, miquel and biel.

\begin{abstract}
Vision based motion estimation algorithms are widely used in ground-based and aerial robotics. Combined with inertial measurement units, they have proven to be a precise and low-cost sensor for velocity and pose estimation. In this paper we show that stereo vision based odometry can be used by autonomous underwater vehicles that navigate close to the seabed for velocity and incremental pose estimation in small areas. We present the integration of two different stereo visual odometry algorithms into an AUV and experiments carried out in laboratory and harbour conditions comparing vision based pose estimates with ground truth.
\end{abstract}

@@

The latest advances in Autonomous Underwater Vehicles (AUVs) demand accurate estimations of the vehicle's pose and velocity. The tasks to be performed by such robots are increasingly complex and, in most cases it is needed to precisely position the underwater vehicle to a certain location in order to execute the desired task. Navigation systems based on wirelessly transmitted data, such as the Global Positioning System (GPS) or mobile network positioning, can not be used in underwater environments. One can overcome this issue by introducing artificial landmarks. However the cost and effort for this is high and for small missions it is preferred to rely on positioning using self-contained sensors. One of the most popular positioning sensors for underwater vehicles is the Doppler Velocity Log (DVL) which provides precise velocity and altitude updates. However, it is an expensive device that cannot be integrated in many underwater vehicles.

@@

First, the paper explains related works on visual odometry. Section \ref{visual-odometry-systems} describes the general concepts of visual odometry systems and presents a brief overview of the two analyzed algorithms, followed by the explanation of the evaluation method in Section \ref{experiments} and experiments in different evironments. Finally, the conclusion of the study and future perspectives are presented.

@@

The basic visual odometry algorithm pipeline \cite{Moravec1980} consists of the following steps: First, keypoints (landmarks) are identified in each camera frame and feature descriptors for these points are extracted. Then, the depth for every landmark is estimated using stereo, structure from motion or a separate depth camera. Subsequently, features are matched across time frames and the rigid-body transformation that best aligns the features between frames is estimated. The result of this process is an estimation of camera motion between frames and therefore it is necessary to integrate this data over time to obtain the vehicle's absolute position and orientation.

@@

We created wrappers for both libraries\footnote{See \url{http://www.ros.org/wiki/viso2_ros} and \url{http://www.ros.org/wiki/fovis_ros} for source code and documentation.} to integrate them into the Robot Operating System ROS \cite{Quigley2009}. This simplifies the integration in different robotic vehicles as well as comparison with other integrated sensors.

@@

The features used by libviso2 are simple blob and corner detector masks, resulting in a large amount of interest points of four different classes. Non-minima-/non-maxima-suppression is used for initial feature reduction. In this step, two different thresholds for minima/maxima selection are used resulting in two sets of sparse strong features and dense less strong features. This allows for a multi-stage matching coarse to fine using matches of the strong features to limit the search range for the weaker ones. The used descriptor contains only 16 values of sobel filter responses distributed inside an 11x11 pixel window. The similarity measure for two features is the sum of absolute differences of their descriptors. The computation of this measure is sped up using SIMD instructions.

@@

Apart from the used feature detectors, an important difference of the two algorithms is the selection of the reference frame for motion computation. libviso2 does not have automatic selection of the reference frame and replaces it in each iteration. In contrast, fovis only replaces the reference frame if the number of inlier matches drops below a given threshold. The purpose is to reduce drift that could arise from sensor noise or bad calibration.

@@

The experiments carried out use data gathered during the TRIDENT project\footnote{See \url{http://www.irs.uji.es/trident/}} both in laboratory and harbour conditions using the Girona500 AUV \cite{Ribas2012} as vehicle (Figure \ref{g500}). This vehicle has been designed by the Universitat de Girona and, without change, does not have stereo vision system, but allows the mounting of external systems on the bottom payload area. This featured is used to attach the stereo vision unit developed by our Group in the Universitat de les Illes Balears. This unit is called Fugu-Flexible, or Fugu-F to shorten, which is basically composed of two stereo rigs and a computer system. Each module, hardware and cameras, is placed in an independent sealed case rated for up to 100 meters depth. The main Fugu-F hardware specifications are:

\begin{itemize}
  \item Two Point Grey Bumblebee2 stereo cameras, one with a focal distance of 3.8mm (66\degree HFOV), and the other with 2.5mm (97\degree HFOV). % HFOV in water?
  \item A mini-ITX motherboard from Jetway-NF98-QM57 Express Chipset with an Intel i5 processor at 2.33GHz with 4 cores.
  \item A PCI express card with two firewire ports from Chronos to connect the cameras to the motherboard.
\end{itemize}

@@

In the harbour experiment, a DVL and an AHS have been used as navigation sensors to let the vehicle follow a previously defined trajectory autonomously. In the laboratory experiments, the vehicle has been remotely controlled by a human operator. Table \ref{experiment-details} summarizes the characteristics of the experiments.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the ones of the bottom refer to harbour tests. Linear velocity is shown in left graphics for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

@@

Results for fovis odometer are substantially different in translation error comparing laboratory and harbour conditions. By examining the number of inliers of the two odometers shown in Table \ref{results-runtime} it is easy to see that fovis is significantly less robust in harbour enviroments where the quality of the images is lower (see Figure \ref{sample-images}). Furthermore, unlike libviso2, fovis presents a better performance when estimating the angular motion in harbour conditions. This improvement over libviso2 is due to the method of keyframe selection based on the number of matches introduced by fovis. This technique has a positive impact when the vehicle experiences pure rotational movements, as it can keep the reference frame for longer and therefore reduce the drift significally.

>>>>

\begin{abstract}
Vision based motion estimation algorithms are widely used in ground-based and aerial robotics. Combined with inertial measurement units, they have proven to be a precise and low-cost sensor for velocity and pose estimation. In this paper we show that stereo vision based odometry can be used by autonomous underwater vehicles (AUV) that navigate close to the seabed for velocity and incremental pose estimation in small areas. We present the integration of two different stereo visual odometry algorithms into an AUV and experiments carried out in laboratory and harbour conditions comparing vision based pose estimates with ground truth.
\end{abstract}

@@

The latest advances in Autonomous Underwater Vehicles (AUVs) demand accurate estimations of the vehicle's pose and velocity. The tasks to be performed by such robots are increasingly complex and, in most cases it is needed to precisely position the underwater vehicle to a certain location in order to execute the desired task. Navigation systems based on wirelessly transmitted data, such as the Global Positioning System (GPS) or mobile network positioning, can not be used in underwater environments. One can overcome this issue by introducing artificial landmarks. However the cost and effort for this is high and for small missions it is preferred to rely on positioning using self-contained sensors. One of the most popular positioning sensors for underwater vehicles is the Doppler Velocity Log (DVL) which provides precise velocity and altitude updates. However, it is a big and expensive device that cannot be integrated in many underwater vehicles.

@@

First, the paper explains related work on visual odometry. Section \ref{visual-odometry-systems} describes the general concepts of visual odometry systems and presents a brief overview of the two analyzed algorithms, followed by the explanation of the evaluation method in Section \ref{experiments} and experiments in different evironments. Finally, the conclusion of the study and future perspectives are presented.

@@

The basic visual odometry algorithm pipeline \cite{Moravec1980} consists of the following steps: first, keypoints (landmarks) are identified in each camera frame and feature descriptors for these points are extracted. Then, the depth for every landmark is estimated using stereo, structure from motion or a separate depth camera. Subsequently, features are matched across time frames and the rigid-body transformation that best aligns the features between frames is estimated. The result of this process is an estimation of camera motion between frames and therefore it is necessary to integrate this data over time to obtain the vehicle's absolute position and orientation.

@@

We created wrappers for both libraries\footnote{See \url{http://www.ros.org/wiki/viso2_ros} and \url{http://www.ros.org/wiki/fovis_ros} for source code and documentation.} to integrate them into the \textit{Robot Operating System} ROS \cite{Quigley2009}. This simplifies the integration in different robotic vehicles as well as comparison with other integrated sensors.

@@

The features used by libviso2 are simple blob and corner detector masks, resulting in a large amount of interest points of four different classes. Non-minima-/non-maxima-suppression is used for initial feature reduction. In this step, two different thresholds for minima/maxima selection are used resulting in two sets of sparse strong features and dense less strong features. This allows for a multi-stage matching coarse to fine using matches of the strong features to limit the search range for the weaker ones. The used descriptor contains only 16 values of Sobel filter responses distributed inside an 11x11 pixel window. The similarity measure for two features is the sum of absolute differences of their descriptors. The computation of this measure is sped up using SIMD instructions.

@@

Apart from the used feature detectors, an important difference of the two algorithms is the selection of the reference frame for motion computation. libviso2 does not have automatic selection of the reference frame and replaces it in each iteration. In contrast, fovis only replaces the reference frame if the number of inlier matches drops below a given threshold. Its purpose is to reduce the drift that could arise from sensor noise or bad calibration.

@@

The experiments carried out use data gathered during the TRIDENT project\footnote{See \url{http://www.irs.uji.es/trident/}} both in laboratory and harbour conditions using the Girona500 AUV \cite{Ribas2012} as vehicle (Figure \ref{g500}). This vehicle has been designed by the Universitat de Girona and, without change, does not have stereo vision system, but allows the mounting of external systems on the bottom payload area. This feature is used to attach the stereo vision unit developed by our group in the Universitat de les Illes Balears. This unit is called Fugu-Flexible which is basically composed of two stereo rigs and a computer system. Each module, hardware and cameras, is placed in an independent sealed case rated for up to 100 meters depth. The main Fugu-Flexible hardware specifications are:

\begin{itemize}
  \item Two stereo cameras, one with a focal distance of 3.8mm (66\degree HFOV), and the other with 2.5mm (97\degree HFOV). % HFOV in water?
  \item A motherboard with an Intel i5 processor at 2.33GHz with 4 cores.
  \item A PCI express card with two firewire ports to connect the cameras to the motherboard.
\end{itemize}

@@

In the harbour experiment, a DVL and an AHS (Airborne Hyperspectral Scanner) have been used as navigation sensors to let the vehicle follow a previously defined trajectory autonomously. In the laboratory experiments, the vehicle has been remotely controlled by a human operator. Table \ref{experiment-details} summarizes the characteristics of the experiments.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the bottom ones refer to harbour tests. Linear velocity is shown on the left for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

@@

Results for fovis odometer are substantially different in translation error comparing laboratory and harbour conditions. By examining the number of inliers of the two odometers shown in Table \ref{results-runtime} it is easy to see that fovis is significantly less robust in harbour enviroments where the quality of the images is lower and the seabed has a poor texture (see Figure \ref{sample-images}). Furthermore, unlike libviso2, fovis presents a better performance when estimating the angular motion in harbour conditions. This improvement over libviso2 is due to the method of keyframe selection based on the number of matches introduced by fovis. This technique has a positive impact when the vehicle experiences pure rotational movements, as it can keep the reference frame for longer and therefore reduce the drift significally.
